TrueSkill Through Time:\\ the \proglang{Python} and \proglang{R} Packages

### 2020-08-20 article.tex
TTT: corrigo intro, extiendo a filtering-smoothing

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{TrueSkill Through Time: the \proglang{Python} Package}
\Plaintitle{TrueSkill Through Time: the Python Package}
\Shorttitle{TrueSkill Through Time: the \proglang{Python} Package}

@@

\Keywords{TrueSkill, ranking, skill, education, sports, games, Bayesian inference, \proglang{Python}}
\Plainkeywords{TrueSkill, ranking, skill, education, sports, games,  Bayesian inference, Python}

@@

\todo[inline]{Me gusta que intentas poner un poco de intro, pero no me parece que quede bien empezar con preguntas}
\en{What are the skills of the people?}
\es{¿Cu\'ales son las habilidades de las personas?}
\en{How difficult are the challenges they face?}
\es{¿Cu\'an dif\'iciles son los desaf\'ios que enfrentan?}
\en{How are their learning curves over time?}
\es{¿C\'omo son sus curvas de aprendizaje en el tiempo?}
\en{What is the overall pattern and how variable is it?}
\es{¿Cu\'al es el patr\'on general y cu\'an variable es?}
\en{Knowing people's abilities is crucial in the area of education and sports, among others.}
\es{Conocer las habilidades de las personas es crucial en el \'area de la educaci\'on y los deportes, entre otros.}
\en{In the context of interventions, it makes it possible to avoid exposing agents to imbalanced challenges.}
\es{En las contexto de intervenciones, permite evitar exponer a los agentes a desaf\'ios desproporcionados.}
\en{For research purposes, the ratings can be made available to subjects and the general public, stimulating interest and engagement.}
\es{Con fines de investigaci\'on, las calificaciones pueden ponerse a disposici\'on de los sujetos y del p\'ublico general, estimulando as\'i el inter\'es y la participaci\'on.}
\en{However, since it is a complex variable, it is generally impossible to measure it directly (i.e. hidden variable).}
\es{Sin embargo, al ser una variable compleja, generalmente es imposible medirla directamente (aka variable oculta).}
\en{In practice, the best we can do is estimate it from its direct observable consequences.}
\es{En la pr\'actica, lo mejor que podemos hacer es estimarla a partir de sus consecuencias observables m\'as directas.}
\en{The most widely used sources of information in education and sports are the outcomes of problem solving and competition between opponents, respectively.}
\es{Las fuentes de informaci\'on m\'as utilizadas en educaci\'on y deportes provienen del producto de resoluci\'on de problemas y de competencia entre oponentes, respectivamente.}
\en{The naive solution, frequency of positive results, is not a good indicator essentially because it also depends on the difficulty of the challenges.}
\es{La soluci\'on ingenua, frecuencia de resultados positivos, no es un buen indicador fundamentalmente porque depende tambi\'en de la dificultad de los desaf\'ios.}
\en{For that reason, all conventional models are underpinned by pairwise comparisons.}
\es{Por esa raz\'on, todos los modelos convencionales se basan en \emph{pairwise comparisons}.}
\en{Since the first generative models, proposed almost a century ago by~\cite{thurstone1927-comparativeJugement} and~\cite{zermelo1929-translation}, it is assumed that the probability of an observed result, $r$, depends on the difference in strength, $d$, between the agent and their target.}
\es{Ya los primeros modelos generativos, propuestos hace casi un siglo por~\cite{thurstone1927-comparativeJugement} y~\cite{zermelo1929-translation}, suponen que la probabilidad de un resultado observado, $r$, depende de la diferencia de escala, $d$, entre el agente y su objetivo.}
\begin{equation} \label{eq:pairwise-comparision}
p(\,r \,|\, d_{ij}\,) 
\en{The field was revitalized after 20 years by the publications of~\cite{bradley1952-pairedComparisions} and~\cite{mosteller1951-remarks}.}
\es{El campo revitaliz\'o 20 a\~nos despu\'es gracias a las publicaciones de \cite{bradley1952-pairedComparisions} y~\cite{mosteller1951-remarks}.}
\en{Nowadays, the most widely used skill estimators are those proposed by~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, and some of the Item-Response Theory variants compiled by~\cite{vanDerLinden2016-handbookIRT_1} and~\cite{fox2010-bayesianIRT}.}
\es{Actualmente, los estimadores de habilidades m\'as utilizados son los propuestos por~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, y algunas de las variantes de la\emph{Item-Response Theory} recopiladas por~\cite{vanDerLinden2016-handbookIRT_1} y~\cite{fox2010-bayesianIRT}.}
\en{Remarkably, they all share the same underlying generative model.}
\es{Notablemente, todos comparten el mismo modelo generativo subyacente.}
\en{The main difference in respect to their representations is limited to the choice of one of two very similar cumulative probability distributions to define the equation~\ref{eq:pairwise-comparision}: the logistic function or the cumulative Gaussian one.}
\es{Respecto de su representaci\'on, la principal diferencia se limita a la elecci\'on de una de dos distribuciones de probabilidad acumulativa, muy similares entre s\'i, para definir la ecuaci\'on~\ref{eq:pairwise-comparision}: la funci\'on log\'istica o la Gaussiana acumulativa.}
\en{The most substantive differences are methodological differences: alternative methods for solving the shared probabilistic model.}
\es{Las diferencias sustantivas son diferencias metodol\'ogicas: m\'etodos alternativos para resoluci\'on de el modelo probabil\'istico.}

@@

    \edge {d} {r};
    \edge {p1,p2} {d};
    \edge {s1} {p1};
    \edge {s2} {p2};
    %\node[invisible, right=of p2, xshift=4.35cm] (s-dist) {};
}
     \caption{
     \en{Clasical generative model in which hidden skills cause the observable results through the difference of random performances.}
    \es{Modelo generativo cl\'asico en la que las habilidades ocultas causan los resultados observables a trav\'es de la diferencia de rendimientos aleatorios.}
    %
    \en{The graphical representation defines a joint probability distribution.}
    \es{La representaci\'on gr\'afica define una distribuci\'on de probabilidad conjunta.}
    %
    \en{Black dots denote constant values.}
    \es{Los puntos negros denotan valores constantes.}
    %
    \en{The big circles and squares indicate continuous and discrete random variables respectively.}
    \es{Los c\'irculos y cuadrados indican variables aleatorias continuas y discretas respectivamente.} 
    %
    \en{Observable variables are painted gray and hidden ones are left blank.}
    \es{Las variables observables se pintan de gris y las variable ocultas se dejan en blanco.}
    %
    \en{The arrows represent causal relationships and the functions define them.}
    \es{Las flechas representan relaciones causales y las funciones las definen.}
    }
    \label{fig:generative_model}
\label{modelo}
\end{figure}

@@

\en{Although agents may have a specific skill at any given time ($s$), the performance ($p$) they exhibit in an event depends on an undetermined number of unknown factors, which are modeled as random noise ($\beta$), $\N(p\,|\,s,\beta^2)$.}
\es{Si bien los agentes pueden tener una habilidad espec\'ifica en cada momento ($s$), el desempe\~no ($p$) que exhiben en cada evento depende de una cantidad indeterminada de factores desconocidos, los que se modelan como ruido aleatorio ($\beta$), $\N(p\,|\,s,\beta^2)$.}

@@

\en{According to the model, the observed results are a direct consequence of the difference of those hidden performances, $d_{ij}=p_i - p_j$.}
\es{Según el modelo, los resultados son una consecuencias directa de la diferencia de esos rendimientos ocultos, $d_{ij}=p_i - p_j$.}

@@

\en{An agent gets a positive result when he outperforms his opponent, $r_{ij} = (d_{ij} > 0)$.}
\es{Un agente obtiene un resultado positivo cuando supera el rendimiento de su adversario, $r_{ij} = (d_{ij} > 0)$.}

@@

 \Delta = \underbrace{y_{ij}}_{\hfrac{\text{Direction}}{\text{(Sign)}}}\underbrace{\left(1-p(\,r_{ij}\,|\,s_i,s_j\,)\right)}_{\hfrac{\text{Magnitud}}{\text{(Surprise)}}}

@@

 s_i^{\text{new}} = s_i^{\text{old}} + \Delta

@@

\en{That is the role played by the famous K-factor used by FIDE.}
\es{Ese es rol que desempe\~na el famoso K-factor usado por la FIDE.}

@@

\en{Probability Theory provides us with a consistent framework for quantifying the uncertainty of our beliefs~\cite{bishop2006-PRML}.}
\es{La Teor\'ia de la Probabilidad nos proporciona un marco consistente para cuantificar la incertidumbre de nuestras creencias~\citep{bishop2006-PRML}.}
\en{The use of probability theory to handle uncertainty, known as Bayesian inference, is now widely adopted in the area of artificial intelligence and machine learning primarily because it has proved to be a extremely useful tool in practice.}
\es{El uso de la teor\'ia de la probabilidad para manejar la incertidumbre, conocida como inferencia bayesiana, est\'a hoy adopta de manera generalizada en el \'ambito de la inteligencia artificial y el aprendizaje autom\'atico fundamentalmente porque ha mostrado ser un instrumento sumamente útil en la pr\'actica.}
\en{It is not however an ad-hoc choice.}
\es{No es sin embargo una elecci\'on ad-hoc.}
\en{Any system that represent degrees of belief by a numerical value, consistent with classical deductive logic and a set of axioms encoding commonsense reasoning, will necessarily be isomorphic to probability theory~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
\es{Cualquier sistema que represente grados de creencia mediante un valor num\'erico, consistente con la l\'ogica deductiva cl\'asica y un conjunto de axiomas que codifican el razonamiento de sentido común,  ser\'a necesariamente isom\'orfico a la teor\'ia de la probabilidad~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
\en{And all of the probabilistic inference, no matter how complex, can be solved in terms of two simple equations: the~\ref{eq:sum_rule} and the~\ref{eq:product_rule}.}
\es{Y toda inferencia probabil\'istica, no importa cuan compleja ella sea, puede ser resuelta a trav\'es de dos simples ecuaciones: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}.}

@@

 p(x) = \sum_{y \in Y} p(x,y)

@@

\en{This formula allows us to optimally update our beliefs about a set of hypotheses, given a model and the data.}
\es{Esta expresi\'on nos permite actualizar de forma \'optima nuestras creencia sobre ciertas hip\'otesis, dado un modelo y los datos.}

@@

 \underbrace{p(\text{\en{Hypotesis}\es{Hip\'otesis}}|\text{\en{Data}\es{Datos}})}_{\text{Posterior\es{i}}} = \frac{\overbrace{p(\text{\en{Data}\es{Datos}}|\text{\en{Hypotesis}\es{Hip\'otesis}})}^{\text{\en{Likelihood}\es{Verosimilitud}}}\overbrace{p(\text{\en{Hypotesis}\es{Hip\'otesis}})}^{\text{Prior\es{i}}}}{\underbrace{p(\text{\en{Data}\es{Datos}})}_{\text{Evidenc\en{e}\es{ia}}}}

@@

\en{where the role played by the model is left implicit.}
\es{donde el lugar que ocupa el modelo queda impl\'icito.}

@@

% Nuevo P\'arrafo

\en{In our case, the hypotheses are the people's ability, the data is given by the observed results, and the model is defined by the figure~\ref{fig:generative_model}.}
\es{En nuestro caso, las hip\'otesis son las habilidades de las personas, los datos son los resultados y el modelo est\'a definido por la figure~\ref{fig:generative_model}.}

@@

\en{With a probability distribution that represents our prior beliefs about people's abilities, we can obtain the exact posterior belief distribution by the use of the sum and product rules.}
\es{Con una distribuci\'on de probabilidad que represente nuestras creencias a priori sobre las habilidad de las personas, podemos obtener la distribuci\'on de creencias a posterori exacta usando tan solo las reglas de la suma y el producto.}

@@

\en{}
\es{En el caso ganador}

@@

\begin{figure}[t!]

@@

\en{The posterior distribution is obtained as the product of the prior and the likelihood.}
\es{La distribuci\'on a posteriori se obtiene como el producto de la distribuci\'on a priori y la verosimilitud.}

@@

\en{It is interesting to note that likelihood works as a filter for prior beliefs.}
\es{Es interesante observar que la verosimilitud funciona como filtro de las creencias a priori.}
\en{In regions of very high skill, where an observed winning result would not generate any surprises, the a posteriori distribution stays next to the prior.}
\es{En las regiones de habilidad muy alta, donde un resultado ganador observado no generar\'ia ninguna sorpresa, la distribuci\'on a posteriori se queda junto al prior.}
\en{In regions of very low skill, where an observed winning result would be very surprising, the subsequent one loses almost all its density.}
\es{En las regiones de habilidad muy bajas, donde un resultado observado ganador ser\'ia muy sorprendente, la distribuci\'on a posteriori pierde casi toda su densidad.}
\en{It is important to stress that the posterior, although similar, is not a Gaussian.}
\es{Es importante remarcar que la posterior, aunque se parece, no es una Gaussiana.}

@@

@@

@@

%\en{Figure~\ref{fig:posterior_win} shows, in graphical terms, the updating procedure performed in equation~\ref{eq:posterior_win}.}
%\es{La figura~\ref{fig:posterior_win} muestra, en términos gr\'aficos, el procedimiento de actualizaci\'on que se realiza en la ecuaci\'on~\ref{eq:posterior_win}.}
\en{The posterior distribution is obtained as the product of the prior and the likelihood.}
\es{La distribuci\'on a posteriori se obtiene como el producto de la distribuci\'on a priori y la verosimilitud.}
\en{It is interesting to note that likelihood works as a filter for prior beliefs (Figure~\ref{fig:posterior_win}).}
\es{Es interesante observar que la verosimilitud funciona como filtro de las creencias a priori (figura~\ref{fig:posterior_win}).}
\en{It is important to stress that the posterior, although similar, is not a Gaussian.}
\es{Es importante remarcar que la posterior, aunque se parece, no es una Gaussiana.}

% Parrafo nuevo

\en{Obtaining the Gaussian that best approximates it would allow us to extend the inference iteratively in time through the use of the posteriors as priors of the following observations.}
\es{Conseguir la Gaussiana que mejor la aproxima nos permitir\'ia extender la inferencia iterativamente en el tiempo a tavés del uso de las posteriors como priors de las siguientes observaciones.}
%
\en{One of the main contributions of~\cite{herbrich2006-trueskill} was to offer an efficient method to find that Gaussian.}
\es{Uno de los principales aportes de~\cite{herbrich2006-trueskill} fue ofrecer un método eficiente para calcular esa Gaussiana.}

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{TrueSkill Through Time:\\ the \proglang{Python} and \proglang{R} Packages}
\Plaintitle{TrueSkill Through Time:\\ the Python and R Packages}
\Shorttitle{TrueSkill Through Time:\\ the \proglang{Python} and \proglang{R} Packages}

@@

\Keywords{TrueSkill, ranking, skill, education, sports, games, Bayesian inference, \proglang{Python}, \proglang{R}}
\Plainkeywords{TrueSkill, ranking, skill, education, sports, games,  Bayesian inference, Python, R}

@@

% \en{What are the skills of the people?}
% \es{¿Cu\'ales son las habilidades de las personas?}
% %   
% \en{How difficult are the challenges they face?}
% \es{¿Cu\'an dif\'iciles son los desaf\'ios que enfrentan?}
% %
% \en{How are their learning curves over time?}
% \es{¿C\'omo son sus curvas de aprendizaje en el tiempo?}
% %
% \en{What is the overall pattern and how variable is it?}
% \es{¿Cu\'al es el patr\'on general y cu\'an variable es?}
% %
\en{Knowing people's abilities is crucial in the area of education and sports.}
\es{Conocer las habilidades de las personas es crucial en el \'area de la educaci\'on y los deportes.}
%
\en{In the context of interventions, it makes it possible to avoid exposing agents to imbalanced challenges.}
\es{En las contexto de intervenciones, permite evitar exponer a los agentes a desaf\'ios desproporcionados.}
%
\en{For research purposes, the ratings can be made available to subjects and the general public, stimulating interest and engagement.}
\es{Con fines de investigaci\'on, las calificaciones pueden ponerse a disposici\'on de los sujetos y del p\'ublico general, estimulando as\'i el inter\'es y la participaci\'on.}
%
\en{However, since it is a complex variable, it is generally impossible to measure it directly (i.e. hidden variable).}
\es{Sin embargo, al ser una variable compleja, generalmente es imposible medirla directamente (aka variable oculta).}
%
\en{In practice, the best we can do is estimate it from its direct observable consequences.}
\es{En la pr\'actica, lo mejor que podemos hacer es estimarla a partir de sus consecuencias observables m\'as directas.}
%
\en{The most widely used sources of information in education and sports are the outcomes of problem solving and competition between opponents, respectively.}
\es{Las fuentes de informaci\'on m\'as utilizadas en educaci\'on y deportes provienen del producto de resoluci\'on de problemas y de competencia entre oponentes, respectivamente.}
%
\en{The naive solution, frequency of positive results, is not a good indicator essentially because it also depends on the difficulty of the challenges.}
\es{La soluci\'on ingenua, frecuencia de resultados positivos, no es un buen indicador fundamentalmente porque depende tambi\'en de la dificultad de los desaf\'ios.}
%
\en{For that reason, all conventional models are underpinned by pairwise comparisons.}
\es{Por esa raz\'on, todos los modelos convencionales se basan en \emph{pairwise comparisons}.}
%
\en{Since the first generative models, proposed almost a century ago by~\cite{thurstone1927-comparativeJugement} and~\cite{zermelo1929-translation}, it is assumed that the probability of an observed result, $r$, depends on the difference in strength, $d$, between the agent and their target, $P(\, r \,|\, d \,)$.}
\es{Ya los primeros modelos generativos, propuestos hace casi un siglo por~\cite{thurstone1927-comparativeJugement} y~\cite{zermelo1929-translation}, suponen que la probabilidad de un resultado observado, $r$, depende de la diferencia de escala, $d$, entre el agente y su objetivo, $P(\, r \,|\, d \,)$.}
%
\en{The field was revitalized after 20 years by the publications of~\cite{bradley1952-pairedComparisions} and~\cite{mosteller1951-remarks}.}
\es{El campo revitaliz\'o 20 a\~nos despu\'es gracias a las publicaciones de \cite{bradley1952-pairedComparisions} y~\cite{mosteller1951-remarks}.}
%

% Parrafo

%
\en{Nowadays, the most widely used skill estimators are those proposed by~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, and some of the Item-Response Theory variants compiled by~\cite{vanDerLinden2016-handbookIRT_1} and~\cite{fox2010-bayesianIRT}.}
\es{Actualmente, los estimadores de habilidades m\'as utilizados son los propuestos por~\cite{elo1961-uscf},~\cite{glikman_gliko_2},~\cite{herbrich2006-trueskill}, y algunas de las variantes de la\emph{Item-Response Theory} recopiladas por~\cite{vanDerLinden2016-handbookIRT_1} y~\cite{fox2010-bayesianIRT}.}
%
\en{Remarkably, they all share the same underlying generative model.}
\es{Notablemente, todos comparten el mismo modelo generativo subyacente.}
%
\en{The main difference in respect to their representations is limited to the choice between two very similar cumulative probability distributions to define the likelihood $P(\, r \,|\, d \,)$: the logistic or the Gaussian cumulative function.}
\es{La principal diferencia respecto a sus representación se limita a la elección entre dos distribuciones de probabilidad acumulativas muy similares para definir la verosimilitud $P(\, r \,|\, d \,)$: la función logística o la Gaussiana acumulativa.}
%
\en{The most substantive differences are methodological differences: alternative methods for solving the shared probabilistic model.}
\es{Las diferencias sustantivas son diferencias metodol\'ogicas: m\'etodos alternativos para resoluci\'on de el modelo probabil\'istico.}

@@

    \edge {d} {r};
    \edge {p1,p2} {d};
    \edge {s1} {p1};
    \edge {s2} {p2};
    %\node[invisible, right=of p2, xshift=4.35cm] (s-dist) {};
}
     \caption{
     \en{Clasical generative model in which hidden skills cause the observable results mediated by the difference of hidden performances.}
    \es{Modelo generativo cl\'asico en el que las habilidades ocultas causan los resultados observables a trav\'es de la diferencia de rendimientos ocultos.}
    %
    \en{At each event, agents exhibit a random performance around their true skill, $p \sim \N(s,\beta^2)$.}
    \es{En cada evento, los agentes exhiben una rendimiento aleatorio centrado en su verdadera habilidad, $p \sim \N(s,\beta^2)$.}
    %
    \en{The one with the highest performance wins, $r_{ij} = p_i > p_j$.}
    \es{Gana quien haya obtenido mayor rendimiento, $r_{ij} = p_i > p_j$.}
    %
    \en{The graphical representation defines a joint probability distribution.}
    \es{La representaci\'on gr\'afica define una distribuci\'on de probabilidad conjunta.}
    %
    \en{By observing a result we can deduce what happened to the hidden variables, and update our hypotheses.}
    \es{Al observar un resultado podemos deducir lo que ocurrió con las variables ocultas, y actulizar nuestras hipótesis.}
    %
%     \en{Black dots denote constant values.}
%     \es{Los puntos negros denotan valores constantes.}
%     %
%     \en{The big circles and squares indicate continuous and discrete random variables respectively.}
%     \es{Los c\'irculos y cuadrados indican variables aleatorias continuas y discretas respectivamente.} 
%     %
%     \en{Observable variables are painted gray and hidden ones are left blank.}
%     \es{Las variables observables se pintan de gris y las variable ocultas se dejan en blanco.}
%     %
%     \en{The arrows represent causal relationships and the functions define them.}
%     \es{Las flechas representan relaciones causales y las funciones las definen.}
    }
    \label{fig:generative_model}
\label{modelo}
\end{figure}

@@

\en{Although agents may have a specific skill ($s$), the performance ($p$) they exhibit in an event depends on an undetermined number of unknown factors, which are modeled with a noise ($\beta$), $\N(p\,|\,s,\beta^2)$.}
\es{Si bien los agentes tienen una habilidad ($s$), el desempe\~no ($p$) que exhiben en cada evento depende de una cantidad indeterminada de factores desconocidos, que se modelan con un ruido ($\beta$), $\N(p\,|\,s,\beta^2)$.}

@@

\en{The observed results are a direct consequence of the difference of those hidden performances, $d_{ij}=p_i - p_j$.}
\es{Los resultados son una consecuencias directa de la diferencia de esos rendimientos ocultos, $d_{ij}=p_i - p_j$.}

@@

\en{The one with the highest performance wins, $r_{ij} = (d_{ij} > 0)$.}
\es{Gana quien haya obtenido mayor rendimiento, $r_{ij} = (d_{ij} > 0)$.}

@@

 \Delta_i = \underbrace{y_{ij}}_{\hfrac{\text{Direction}}{\text{(Sign)}}}\underbrace{\left(1-P(\,r_{ij}\,|\,s_i,s_j\,)\right)}_{\hfrac{\text{Magnitud}}{\text{(Surprise)}}}

@@

 s_i^{\text{new}} = s_i^{\text{old}} + \Delta_i

@@

\en{That is the role played by the K-factor used by the FIDE,  $s_i^{\text{new}} = s_i^{\text{old}} + \Delta_i K_i$.}
\es{Ese es rol que desempe\~na el K-factor usado por la FIDE, $s_i^{\text{new}} = s_i^{\text{old}} + \Delta_i K_i$.}

@@

\en{Probability Theory provides us with a consistent framework for quantifying the uncertainty of our beliefs~\cite{bishop2006-PRML}.}
\es{La Teor\'ia de la Probabilidad nos proporciona un marco consistente para cuantificar la incertidumbre de nuestras creencias~\citep{bishop2006-PRML}.}
%
\en{The use of probability theory to handle uncertainty, known as Bayesian inference, is now widely adopted in the area of artificial intelligence and machine learning primarily because it has proved to be a extremely useful tool in practice.}
\es{El uso de la teor\'ia de la probabilidad para manejar la incertidumbre, conocida como inferencia bayesiana, est\'a hoy adopta de manera generalizada en el \'ambito de la inteligencia artificial y el aprendizaje autom\'atico fundamentalmente porque ha mostrado ser un instrumento sumamente útil en la pr\'actica.}
%
\en{It is not however an ad-hoc choice.}
\es{No es sin embargo una elecci\'on ad-hoc.}
%
\en{Any system that represent degrees of belief by a numerical value, consistent with classical deductive logic and a set of axioms encoding commonsense reasoning, will necessarily be isomorphic to probability theory~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
\es{Cualquier sistema que represente grados de creencia mediante un valor num\'erico, consistente con la l\'ogica deductiva cl\'asica y un conjunto de axiomas que codifican el razonamiento de sentido común,  ser\'a necesariamente isom\'orfico a la teor\'ia de la probabilidad~\citep{cox1946-reasonableExpectation,vanHorn2003-guideCoxTheorem}.}
%
\en{All of the probabilistic inference, no matter how complex, can be solved in terms of two simple equations: the~\ref{eq:sum_rule} and the~\ref{eq:product_rule}.}
\es{Toda inferencia probabil\'istica, no importa cuan compleja ella sea, puede ser resuelta a trav\'es de dos simples ecuaciones: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}.}

@@

 P(x) = \sum_{y \in Y} P(x,y) \ \ \ \ \ \text{or} \ \ \ \ \ p(x) = \int p(x,y) \, dy

@@

\en{This formula allows us to optimally update our beliefs about space hypotheses, given a model and the data.}
\es{Esta expresi\'on nos permite actualizar de forma \'optima nuestras creencia sobre el espacio de hip\'otesis, dado un modelo y los datos.}

@@

 \underbrace{p(\overbrace{\text{\en{Skills}\es{Habilidades}}}^{\text{\en{Hidden}\es{Oculta}}}|\overbrace{\text{\en{Result}\es{Resultado}}}^{\text{\en{Observed}\es{Observado}}}, \text{\en{Model}\es{Modelo}})}_{\text{Posterior\es{i}}} = \frac{\overbrace{P(\,\text{\en{Result}\es{Resultado}}\,|\,\text{\en{Skills}\es{Habilidades}}\,,\text{\en{Model}\es{Modelo}})}^{\text{\en{Likelihood}\es{Verosimilitud}}}\overbrace{p(\text{\en{Skills}\es{Habilidades}})}^{\text{Prior\es{i}}}}{\underbrace{P(\text{\en{Result}\es{Resultado}}\,|\,\text{\en{Model}\es{Modelo}})}_{\text{Evidenc\en{e}\es{ia}}}}

@@

\en{Where the hypotheses are the hidden skills, the data is given by the observed results, and the model is defined by the figure~\ref{fig:generative_model}.}
\es{Donde las hip\'otesis son las habilidades ocultas, los datos son los resultados y el modelo est\'a definido por la figure~\ref{fig:generative_model}.}

@@

\en{The probability distributions that represent our previous beliefs about people's abilities can be updated by the use of the sum and product rules.}
\es{Las distribuciones de probabilidad que representen nuestras creencias previas sobre las habilidades de las personas se pueden actualizar mediante el uso de las reglas de la suma y el producto.}

@@

\en{Using a Gaussian prior, let's see the exact belief update for the winning case,}
\es{Usando un prior Gaussiano, veamos la actualizaci\'on de creencias exacta para el caso ganador,}

@@

@@

\en{A proprotional posterior distribution is obtained by the product of the prior and the likelihood.}
\es{Una distribuci\'on proporcional al posterior se obtiene mediante el producto de la distribuci\'on a priori y la verosimilitud.}
\en{Figure~\ref{fig:posterior_win} shows, in graphical terms, the updating procedure performed in equation~\ref{eq:posterior_win}.}
\es{La figura~\ref{fig:posterior_win} muestra, en términos gr\'aficos, el procedimiento de actualizaci\'on que se realiza en la ecuaci\'on~\ref{eq:posterior_win}.}
\begin{figure}[h!]

@@

\en{The proprotional posterior distribution is obtained as the product of the prior and the likelihood.}
\es{La distribuci\'on a posteriori proporcional se obtiene como el producto de la distribuci\'on a priori y la verosimilitud.}
\en{The evidence is equivalent to its integration.}
\es{La evidencia es equivalente a su integración.}

@@

@@

\en{It is interesting to note how the likelihood works as a filter for the prior beliefs.}
\es{Es interesante notar cómo la verosimilitud funciona como filtro de las creencias a priori.}

@@

\en{In the regions of very high ability hypothesis, where a winning result would not generate any surprise, the proportional posterior distribution remains unaltered with respect to the prior.}
\es{En las regiones de hipótesis de muy alta habilidad, donde un resultado ganador no generaría sorpresa alguna, la distribución a posteriori proporcional no sufre alteración respecto del prior.}

@@

\en{In the regions of very low ability hypothesis, where a winning result would be very surprising, the proportional posterior distribution loses almost all its mass.}
\es{En las regiones de hipótesis de muy baja habilidad, donde un resultado ganador sería muy sorprendente, la distribución a posteriori proprocional pierde casi toda su masa.}

% Parrafo

\textbf{Ahora, como extendemos la inferencia?}
%
\en{}
\es{}
%
The simplest approach to modelling a sequence of observations is to treat them as independent, corresponding to a set of graph without links between them.
%
\begin{figure}[h!]
  \centering
  \scalebox{.9}{
    \tikz{ %
      \node[latent] (s1) {} ;
      \node[const, above=of s1 ] (ns1) {$s_i^1$} ;
      \node[latent, below=of s1] (p1) {};
      \node[const, right=of p1 ] (np1) {$p_i^1$} ;

      \node[latent, right=of s1] (s2) {} ;
      \node[const, above=of s2 ] (ns2) {$s_i^2$} ;
      \node[latent, below=of s2] (p2) {};
      \node[const, right=of p2 ] (np2) {$p_i^2$} ;

      \node[const, right=of s2, xshift=0.5cm] (s3) {\ $\dots$} ;

      \node[latent, right=of s3, xshift=-0.4cm] (s1n) {} ;
      \node[const, above=of s1n ] (ns1n) {$s_i^{n-1}$} ;
      \node[latent, below=of s1n] (p1n) {};
      \node[const, right=of p1n ] (np1n) {$p_i^{n-1}$} ;

      \node[latent, right=of s1n] (sn) {} ;
      \node[const, above=of sn ] (nsn) {$s_i^n$} ;
      \node[latent, below=of sn] (pn) {};
      \node[const, right=of pn ] (npn) {$p_i^n$} ;

      \edge {s1} {p1};
      \edge {s2} {p2};
      \edge {s1n} {p1n};
      \edge {sn} {pn};
      }  
  }
  \caption{The \emph{filter} approach to modelling a sequence of observations treat each graphical model as independent, without links.}
  \label{fig:filter}
\end{figure}

%Parrafo

\textbf{En este approach generalemnte se usa el posterior con el prior}.
%
Sin embargo, como la habilidad de las personas cambia en el tiempo, se le agrega algo de incertidumbre.
%
\begin{equation}
 \text{Prior\es{i}}_i^t + \N(0,\gamma^2)= \text{Posterior\es{i}}_i^{t-1}
\end{equation}
%
\en{It is important to stress that the exact posterior, although similar, is not a Gaussian distribution.}
\es{Es importante remarcar que la posterior, aunque se parezca, no es una distribución Gaussiana.}
%
\en{Obtaining the Gaussian that best approximates it would allow us to extend the inference iteratively in time through the use of the posteriors as priors of the following observations.}
\es{Conseguir la Gaussiana que mejor la aproxima nos permitir\'ia extender la inferencia iterativamente en el tiempo a tavés del uso de las posteriors como priors de las siguientes observaciones.}
%
\en{One of the main contributions of~\cite{herbrich2006-trueskill} was to offer an efficient method to find that Gaussian.}
\es{Uno de los principales aportes de~\cite{herbrich2006-trueskill} fue ofrecer un método eficiente para calcular esa Gaussiana.}

%
\todo[inline, color=gray!50]{Expectation propagation}

\textbf{TTT}
%
\begin{quotation}
As a skill rating and matchmaking system TrueSkill operates as a filter as discussed above.
However, due to its fully probabilistic formulation it is possible to extend Trueskill to perform smoothing on a time series of player skills.
\end{quotation}
%
\begin{figure}[h!]
  \centering
  \scalebox{.9}{
    \tikz{ %
      \node[latent] (s1) {} ;
      \node[const, above=of s1 ] (ns1) {$s_i^1$} ;
      \node[latent, below=of s1] (p1) {};
      \node[const, right=of p1 ] (np1) {$p_i^1$} ;

      \node[latent, right=of s1] (s2) {} ;
      \node[const, above=of s2 ] (ns2) {$s_i^2$} ;
      \node[latent, below=of s2] (p2) {};
      \node[const, right=of p2 ] (np2) {$p_i^2$} ;

      \node[const, right=of s2, xshift=0.5cm] (s3) {\ $\dots$} ;

      \node[latent, right=of s3, xshift=-0.4cm] (s1n) {} ;
      \node[const, above=of s1n ] (ns1n) {$s_i^{n-1}$} ;
      \node[latent, below=of s1n] (p1n) {};
      \node[const, right=of p1n ] (np1n) {$p_i^{n-1}$} ;

      \node[latent, right=of s1n] (sn) {} ;
      \node[const, above=of sn ] (nsn) {$s_i^n$} ;
      \node[latent, below=of sn] (pn) {};
      \node[const, right=of pn ] (npn) {$p_i^n$} ;

      \edge {s1} {p1,s2};
      \edge {s2} {p2,s3};
      \edge {s3} {s1n};
      \edge {s1n} {p1n, sn};
      \edge {sn} {pn};
      }  
  }
  \caption{The \emph{smoothing} approach treat the sequential data as a unique Markov chain of latent variables. This graphical structure forms the foundation both for the hidden Markov model and for linear dynamical systems.}
  \label{fig:smoothing}
\end{figure}

\begin{quotation}
Although such models are tractable, they are also severely limited. We can obtain a more general framework, while still retaining tractability, by the introduction of latent variables, leading to \emph{state space models}.
%
Here we focus on one important state space models known as \emph{linear dynamical systems} in which the latent variables are Gaussian.
%
In order to motivate the concept of linear dynamical systems, let us consider the following simple problem, which often arises in practical settings.
Suppose we wish to measure the value of an unknown quantity z using a noisy sensor that returns a observation x representing the value of z plus zero-mean Gaussian noise.
our estimate for z by taking lots of measurements and averaging them,
%
Now let’s make the situation more complicated by assuming that we wish to measure a quantity z that is changing over time.
We can take regular measurements of $x$ so that at some point in time we have obtained $x_1 , \dots , x_N$ and we wish to find the corresponding values $z_1 , \dots , z_N$.
If we simply average the measurements, we will introduce a source of error.
%
Intuitively, we could imagine doing a bit better as follows. To estimate the value of $z_N$, we take only the most recent few measurements, say $x_{N −L} , \dots , x_N$ and just average these.
Perhaps we could do even better if we take a weighted average, in which more recent measurements make a greater contribution than less recent ones.
Although this sort of intuitive argument seems plausible, it does not tell us how to form a weighted average, and any sort of hand-crafted weighing is hardly likely to be optimal
%
Fortunately, we can address problems such as this much more systematically by defining a probabilistic model that captures the time evolution and measurement processes and then applying the inference and learning methods developed in earlier chapters.
Here we shall focus on a widely used model known as a linear dynamical system.

\end{quotation}
